---
title: "<center>Predict Exercise Manner by Machine Learning Models</center>"
author: "<center>Shang Ju</center>"
date: "<center>October 28, 2016</center>"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
load('.rdata')
```

## Introduction

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise. 

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, I used data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).

## Data Source

The training data for this project are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

The original work can be found at: 

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. 

-----

## Upload and Explore/Clean Data

Load package `caret`:
```{r}
library(caret)
```

Set seed: (An important step, making sure the result is reproducable) 
```{r}
set.seed(123)
```

Read downloaded raw data:

```{r eval=FALSE}
pml_training <- read.csv('pml-training.csv')
pml_testing <- read.csv('pml-testing.csv')
```

Explore training data:

```{r}
str(pml_training)
```

Training data has 160 variables, including the target variable `classe`, which has five different values:

```{r}
str(training$classe)
```

Explore the testing data:

```{r}
sum(complete.cases(pml_testing))
```

None of the observations in testing data has complete data for all 160 variables. Are all 159 predictors useful at predicting the outcome `classe`? Let's examine the variance of each variable:

```{r}
nzv3 <- nzv(pml_testing,saveMetrics = 1)
nzv3
```

After a deeper examination, it turns out that only 53 predictors are relevant and have some variation at all in the testing data set. So we can filter out the irrelevant variables in both testing and training data:

```{r eval=FALSE}
pml_testing2 <- pml_testing[,c(7:11,37:49,60:68,84:86,102,113:124,140,151:160)]
pml_training2 <- pml_training[,c(7:11,37:49,60:68,84:86,102,113:124,140,151:160)]
```

Note that the testing data does not have the target vaiable `classe` as the last column, but `problem_id` instead:

```{r}
pml_testing2$classe
names(pml_testing2)[length(names(pml_testing2))]
```

## Prepare Data for Modeling

The training dataset has 19622 observations, 53 predictors, and 1 target variable:
```{r}
dim(pml_training2)
```

**Split** training data **70/30** into **training / validation** sets:
```{r eval=FALSE}
inTrain <- createDataPartition(y=pml_training2$classe,p=0.7,list=0)
training <- pml_training2[inTrain,]
validation <- pml_training2[-inTrain,]
testing <- pml_testing2
```

The final dimensions of training/validation/testing sets are:
```{r}
dim(training)
dim(validation)
dim(testing)
```

Note 53 predictors are a lot. Putting all of them into any model would take consirable time to train. So we want to check the correlation between the variables and see if it's possible to use fewer varaibles for modeling.

## Variable Correlation Exploration

Correlation Matrix:
```{r eval=FALSE}
cor(training[,-ncol(training)])
```

Principle Component Analysis:
```{r}
pc <- preProcess(training[,-ncol(training)],method='pca')
pc
```

It seems that PCA could be helpful in modeling for this task.

## Finally: Modeling

First define a function **fit_pred()** to test the accuracy of each model on the validation set:
```{r eval=FALSE}
fit_pred <- function(fit,testSet = validation) {
    pred = predict(fit,testSet);
    confusionMatrix(pred,testSet[['classe']])
}
```

Set seed again:
```{r eval=FALSE}
set.seed(123)
```

Set trainControl method to be **cross-validation** for all models:
```{r eval=FALSE}
ctrl <- trainControl(method="cv", number=10)
```

Build different machine learning models and record the training **time** (bootstrapping / cross-validation). Also record the **accuracy** and **kappa** of each model after applying on the validation set:
```{r eval=FALSE} 
# knn (1081 s / 82 s)
system.time(fit_knn <- train(classe~.,data = training, method = 'knn', trControl=ctrl))
fit_pred(fit_knn) # 0.92 0.90
fit_knn

# knn with pca (239 s / 29 s)
system.time(fit_knn_pca <- train(classe~.,data = training, method = 'knn', preProcess='pca',trControl=ctrl))
fit_pred(fit_knn_pca) # 0.95 0.94
fit_knn_pca

# glm (error, 2-class outcomes only)
system.time(fit_glm_pca <- train(classe~.,data = training, method = 'glm', preProcess='pca'))

# decision tree (26 s / 11 s)
system.time(fit_rpart_pca <- train(classe~.,data = training, method = 'rpart', preProcess='pca', trControl=ctrl))
fit_pred(fit_rpart_pca) # 0.39 0.20
fit_rpart_pca

# decision tree bagged (400 s / 109 s)
system.time(fit_treebag_pca <- train(classe~.,data = training, method = 'treebag', preProcess='pca', trControl=ctrl))
fit_pred(fit_treebag_pca) # 0.96 0.94
fit_treebag_pca

# random forest (2594 s / 838 s)
system.time(fit_rf_pca <- train(classe~.,data = training, method = 'rf', preProcess='pca', trControl=ctrl))
fit_pred(fit_rf_pca) # 0.97 0.97
fit_rf_pca

# random forest cv 5 (383 s)
system.time(fit_rf_pca_cv5 <- train(classe~.,data = training, method = 'rf', preProcess='pca', trControl=trainControl(method='cv',number=5)))
fit_pred(fit_rf_pca_cv5) # 0.97 0.97
fit_rf_pca_cv5

# boosted tree (1188 s / 407 s)
system.time(fit_gbm_pca <- train(classe~.,data = training, method = 'gbm', preProcess='pca',trControl=ctrl))
fit_pred(fit_gbm_pca) # 0.82 0.78
fit_gbm_pca

# linear discriminant analysis (16 s / 5 s)
system.time(fit_lda_pca <- train(classe~.,data = training, method = 'lda', preProcess='pca',trControl=ctrl))
fit_pred(fit_lda_pca) # 0.52 0.40
fit_lda_pca

# naive bayes (913 s / 98 s)
system.time(fit_nb_pca <- train(classe~.,data = training, method = 'nb', preProcess='pca',trControl=ctrl))
fit_pred(fit_nb_pca) # 0.64 0.54
fit_nb_pca

# naive bayes repeated 10 fold cross validation (452 s)
train_control <- trainControl(method="repeatedcv", number=10, repeats=3)
system.time(fit_nb_pca_cv_repeat <- train(classe~.,data = training, method = 'nb', preProcess='pca',trControl=train_control))
fit_pred(fit_nb_pca_cv_repeat) # 0.64 0.54
fit_nb_pca_cv_repeat

# Support Vector Machine with Linear Kernel (669 s / 221 s)
system.time(fit_svm_pca <- train(classe~.,data = training, method = 'svmLinear', preProcess='pca',trControl=ctrl))
fit_pred(fit_svm_pca) # 0.58 0.47
fit_svm_pca

# Support Vector Machine with Polynomial Kernel (7981 s)
system.time(fit_svmpoly_pca <- train(classe~.,data = training, method = 'svmPoly', preProcess='pca',trControl=ctrl))
fit_pred(fit_svmpoly_pca) # 0.98 0.98
fit_svmpoly_pca

# Neural Network (1478 s / 363 s)
system.time(fit_nnet_pca <- train(classe~.,data = training, method = 'nnet', preProcess='pca',trControl=ctrl))
fit_pred(fit_nnet_pca) # 0.60 0.50
fit_nnet_pca
```

Here's the summary of the efficiency/performance of different machine learning models: (Except for the first 'knn' model, all other models were trained using preProcess method '**pca**')

## Model Comparison and Selection

Model                              |  Method in caret  | Time to train (CV) | Accuracy | Kappa
-----------------------------------|-------------------|--------------------|----------|-------
Nearest Neighbor (without PCA)     |    knn            |     82 s           |   92%    |  0.90
Nearest Neighbor                   |    knn            |     29 s           |   95%    |  0.94
Decision Tree                      |   rpart           |     11 s           |   39%    |  0.20
Bagged Tree                        |  treebag          |    109 s           |   96%    |  0.94
Random Forest                      |     rf            |    838 s           |   97%    |  0.97
Random Forest (CV 5-fold)          |     rf            |    383 s           |   97%    |  0.97
Boosted Tree                       |     gbm           |    407 s           |   82%    |  0.78
Linear Discriminant Analysis       |    lda            |      5 s           |   52%    |  0.40
Naive Bayes                        |     nb            |     98 s           |   64%    |  0.54
Naive Bayes (repeated CV)          |     nb            |    452 s           |   64%    |  0.54
Support Vector Machine (Linear Ke) |    svmLinear      |    221 s           |   58%    |  0.47
Support Vector Machine (Poly Ke)   |    svmPoly        |   7981 s           |   98%    |  0.98
Neural Network                     |    nnet           |    363 s           |   60%    |  0.50

### Model Comparison Summary

1. Efficiency and performace vary considerably on this dataset across different models. **'Nearest Neighbor'**, **'Bagged Tree'**, **'Random Forest'**, and **'Support Vector Machine with Polynomial Kernel'** are winner models in terms of prediction accuracy. On the other hand, **'Decision Tree'**,**'Naive Bayes'**,**'Support Vector Machine with Linear Kernel'** and **'Neural Network'** did not predict accurately for this specific task.
2. In terms of speed, **'Decision Tree'** and **'Linear Discriminant Analysis'** run much faster than other models. 

### Model Selection

Considering the models' accuracy and the size of the testing data (20), there's not much more room to improve from model-stacking. So we will proceed to predict the testing data with the most accurate model **'Support Vector Machine with Polynomial Kernel'**. The **expected accurary** of perdicting the 20 cases in testing data should be **100% or 95% (1 case misclassified)**.

## Apply the Winner Model on the Testing Dataset

Finally, we apply the Support Vector Machine model with Polynomial Kernel on the 20 observations in the testing data and below the prediction of their `classe`:

```{r}
predict(fit_svmpoly_pca,testing)
```

And we achieved **95% accuracy** (19 of 20 cases correct).


